---
title: "Week6"
author: "Johanne S. Rejsenhus"
date: "2025-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Good afternoon*  
Today we are going to work with some of the statistical concepts Anna went over yesterday. We will check for normality and implement correlation tests in R. The goal of this markdown is to play around with plotting, transorming and testing assumptions of normality and try to gain an understanding of the concepts. 


**Structure of the markdown:**  
Part 1:
- Visually inspect normality of data
- Test assumptions of normality
- Try standardization
- Perform correlation tests 

Part 2:
- Apply these techniques in a real data example


Start with the first exercise, and then continue in order. Feel free to work together, and see how far you can get.   
The important thing is to learn, not to necessarily do it all!
________________________________________________________________________________________


## Load the packages for today

```{r}
# Install pacman first (only once):
install.packages("pacman")

# Load and install packages:
pacman::p_load(ggplot2, dplyr, tidyr) #So it installs the library if needed + imports it in one step.
```

# Part 1:
For Part one we will walk through the necessary steps of making an analysis of cultural data (except making and fitting models - but that's for next week :)). 

_We would like to investigate whether the personal savings ratio is larger for countries with large disposable income_

## The dataset for the first part
We are gonna work with the df_lcs dataset which is a built-in classic dataset in R, containing economic and demographic indicators for 50 countries, covering the years 1950–1960.

```{r}
df_lcs <- as.data.frame(LifeCycleSavings)
head(df_lcs)
```
The dataset has 50 rows (countries) and 5 numeric variables:

sr – Aggregate personal savings ratio - Average % of disposable personal income that households save.
pop15 – % of population under age 15
pop75 – % of population over age 75
dpi – Real per-capita disposable income
ddpi – Growth rate of dpi


## Check for normality
- Histograms
- QQ-plots
- Statistical normality tests


First we plot the data to visually inspect how the data is distributed - we hope for normal distributions. 
We use 'ggplot' to plot in r and there is so many cool plot you can make to visualize your data. If you would like to try and spice things up, go to: https://r-graph-gallery.com and look around. 

```{r}
# Now we plot the histograms of the savings ratio and the disposable income 

# Savings Ratio
ggplot(df_lcs, aes(x = sr)) + # Here we specify the aesthetics (Which is "just" specifying the x and potentially the y axis.)
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue", color = "black") + # Here we specify the type of plot and that it should be a density plot. 
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Savings Ratio (sr)", x = "Savings Ratio", y = "Density") # And we give it a title and some axis names

#Disposable income
ggplot(df_lcs, aes(x = dpi)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Disposable Income (dpi)", x = "Disposable Income", y = "Density")

```

_Excercise_: Explain what you see, are the distributions normal, if not, describe how they look. 
The plot for sr looks maybe a little normal, maybe
The plot for dpi doesn't really look anything like normal to me.

## QQ plots
The QQ-plot stands for Quantile-Quantile plot. Is plots the thoretical quantiles on the x-axis. That is the quantiles the data would have if it was perfectly normal, and then it plots the actual quantiles on the y-axis. 

If you forgot, quantiles are cut points that divide your data into equal-sized groups. 

```{r}
# Now we plot the QQ-plots of the savings ratio and the disposable income 

ggplot(df_lcs, aes(sample = sr)) + #For qq-plots we specify a 'sample' instead of an x or y axis
  stat_qq(color = "blue") + # Colours 
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Savings Ratio (sr)")

ggplot(df_lcs, aes(sample = dpi)) +
  stat_qq(color = "darkgreen") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Disposable Income (dpi)")

```

_Exercise_: Does this fit with what you observed before when you plotted the histograms? Explain why.
Yeah, i think it fits, because the sr QQ-plot, doesn't look completely right, but not too far off either and the dpi plot looks very much off - It starts way to high and is then not increasing enough for a while.

## Test for normality - Shapiro-wilk test

The Shapiro-Wilk test is a statistical test used to check whether a variable is approximately normally distributed. It is one of the most common tests for normality in R. It test whether there is strong enough evidence to reject the null hypothesis, which is that the data is normal. 

How It Works:
- Null hypothesis (H₀): The data are normally distributed.
- Alternative hypothesis (H₁): The data are not normally distributed.

The test calculates a W statistic that measures how well the data’s distribution matches a normal distribution.


```{r}
shapiro.test(df_lcs$sr)
shapiro.test(df_lcs$dpi)

```

How to interpret the results:
Look at the p-value:
- p > 0.05 → Fail to reject H₀ → Data can be treated as approximately normal.
- p ≤ 0.05 → Reject H₀ → Consider non-parametric methods or transformations.

W statistic:
- W ranges from 0 to 1.
- Values closer to 1 indicate data closer to normal.

_Exercise_: Report the results for both variables, are they normal and why/why not?
For sr, the p-value is 0.58, which is not significant meaning that there is evidence in favor of the data being normal. So it is normal.
For dpi, the p-value is 0.000328 which is signficant meaning that there is evidence an favor of the data not being normal. 
Both of these are in line with what we guessed after the visual inspection.

## Transformation - Lets see if we can force the dpi-variable to become normal. 
We will use log() (logarithmic) and sqrt() (square root) transformation methods and see if it helps.

```{r}

# Log transformation
df_lcs$log_dpi <- log(df_lcs$dpi)

# Sqrt transformation
df_lcs$sqrt_dpi <- sqrt(df_lcs$dpi)

#plot it 
ggplot(df_lcs, aes(x = log_dpi)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Disposable Income (log)", x = "Disposable Income", y = "Density")

ggplot(df_lcs, aes(x = sqrt_dpi)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Disposable Income (sqrt)", x = "Disposable Income", y = "Density")


# test both
shapiro.test(df_lcs$log_dpi)
shapiro.test(df_lcs$sqrt_dpi)

```
_Excercise_: Does any of the transformations make the dpi variable normal?
No not really, we can both see from the graphs and from the Shapiro-Wilk test that it got a bit better, especially the log transformed one, but not enough. The p-value is 0.04, which is just significant unfortunately.

## Standardization 
For this simple correlation analysis, we will skip it, because the results are already interpretable. However, if we wanted to fit a statistical model (e.g., regression) to this data, standardization would be useful. This is because the model coefficients would then represent changes in standard deviations, rather than raw units, making them easier to compare across variables.

This is a method that you will not have to use often, but it is important to know. 

```{r}
# This is how you standardize the data (transform to z-scores)
scaled_data <- scale(df_lcs)
head(scaled_data)

```


_Exercise_: Is standardized data easy to directly interpret? why/why not? (what is a unit now?)
No, it is not easy to interpret, because the units no longer have any real intrinsic meaning attached to them, because the units are now only comparable to other z-scores. 

## Correlation tests
Now, We unfortunately do not have normally distributed data so we should use the spearman correlation test. However, for teaching purposes you will do both and report them to compare the results. 

Pearson:
Measures: Linear relationship
Assumptions: Variables approximately normal, linear, no extreme outliers
Use when: Data are normal and linear

Spearman: 
Measures: Monotonic relationship (rank-based)
Assumptions: None
Use when: Data are skewed, contain outliers, or relationship is monotonic but non-linear

```{r}
# Pearsons correlation - Assumes normality 

cor.test(df_lcs$sr, df_lcs$dpi, method = "pearson")

```
_Excercise_: Report on the results. Whats the confidence interval, the p-value and the correlation and how should they be interpreted?
You can find the APA format of reporting here: https://www.statology.org/how-to-report-pearson-correlation/ :)

The pearsons correlational test found that the percentage of household income that households save (sr) and disposable income per capita (dpi) was not found to be significantly correlated r(48) = 0.22, p = 0.12.
This suggests that there is no difference between how much a family earns, and how much money they save.

The confidence interval was found to be [-0.06 - 0.47] meaning that we are 95% certain that the true mean lies within this interval.

```{r}
# Spearman correlation - does not assume normality

cor.test(df_lcs$sr, df_lcs$dpi, method = "spearman")

```
_Excercise_: Report on the results. Whats the rho and the p-value and how should they be interpreted?
You can find the APA format of reporting Spearmans correlation here: https://www.statology.org/how-to-report-spearman-correlation/ :) 

The sparmans correlational test found that the percentage of household income that households save (sr) and disposable income per capita (dpi) was found to be significantly correlated r(48) = 0.28, p = 0.04.
Contrary to the results from pearsons test, this suggests that there is a signficance correlation between sr and dpi, suggesting that the income of families does have a significant effect on the amount of money they saved.

Okay, so why do we have different result?

_Excercise_: What causes the difference in the results?

It's probably because we used two different tests, a non-parametric and a parametric one.

Well done!! now for part 2


# Part 2:
## Now we try and use our new knowlegde for a mini-project. For this I have made a checklist you can run through

### Step 1: Understand Your Data
- [X] Load the dataset into R  
- [X] Identify which variables are relevant for your analysis  

### Step 2: Explore Distributions
- [X] Plot density histograms for the selected variables - or maybe try some other plots (violinplots, boxplots etc.)
- [X] Create Q-Q plots to visually check normality  
- [X] Run a Shapiro-Wilk test for normality

### Step 3: Standardize Variables (if needed)
- [X] Apply `scale()` to standardize if comparing across variables  

### Step 4: Check Assumptions for Correlation
- [X] If variables look normally distributed → Pearson correlation  
- [X] If variables are not normal or have outliers → Spearman correlation  

### Step 5: Run Correlation Tests
- [ ] Compute Pearson correlation with `cor.test(x, y, method="pearson")`  
- [X] Compute Spearman correlation with `cor.test(x, y, method="spearman")`  
- [X] Interpret results  

### Step 6: Interpret Results
- [X] Report the correlation coefficient (strength & direction of relationship)  
- [X] Report the p-value (evidence for/against significant relationship)  
- [X] Relate findings back to the research question / context of the dataset  


You now have a lot of freedom to find a good question to test, but first lets take a look at the data :)

```{r }
df_states <- as.data.frame(state.x77)
head(df_states)
```

What dataframe are you looking at? Make a note of the column names and their meaning and discuss with your group what an interesting thing to investigate could be. 

Few points of focus:
- Try to set up your analysis like part one 
- Remember to comment your code, so you understand whats happening. 
- Explain whats happening along the way and report on your results

Now the floor is yours ...

```{r}
# Where to begin?
#I will try to look at HS grad (high school graduates) and murder.

ggplot(df_states, aes(x = `HS Grad`)) + #Here, we specifiy that we look at HS grad on the x-axis.
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue", color = "black") + #And here we make the histogram, and specify the bin length and colors.
  geom_density(color = "red", size = 1) + #And we make a density plot.
  labs(title = "Distribution of high school graduation", x = "Percentage that graduated", y = "Density") #We make the titles and labels.

ggplot(df_states, aes(x = Murder)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of murder", x = "Numbrs of murder per 100k capita", y = "Density")
```
None of these look especially normal to me, but out of the two, i would say maybe HS grad looks a tad more normal.


```{r}
#Now for the QQ-Plots
ggplot(df_states, aes(sample = `HS Grad`)) + #He start the plot somewhat the same way as before.
  stat_qq(color = "blue") + # We do the colors
  stat_qq_line(color = "red") + #And we make the QQ line
  labs(title = "Q-Q Plot: HS grad")

ggplot(df_states, aes(sample = Murder)) +
  stat_qq(color = "darkgreen") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Murder")
```
I think the trend is a bit the same as before, murder is not normal at all, and hs grad probably is not either.

```{r}
shapiro.test(df_states$`HS Grad`)
shapiro.test(df_states$Murder)
```

For HS grad, the p-value here is 0.046, which is significant unfortunately, this means that HS grad is not normal.


For murder, the p-value is 0.047, which is also signficant meaning that murder is not normal either.

```{r}
#Now we can try to transform the data in light of the recent findings on normality:
# Log transformation
df_states$log_HS_grad <- log(df_states$`HS Grad`)

# Sqrt transformation
df_states$sqrt_HS_grad <- sqrt(df_states$`HS Grad`)


ggplot(df_states, aes(x = log_HS_grad)) + #Here, we specify that we look at HS grad on the x-axis.
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue", color = "black") + #And here we make the histogram, and specify the bin length and colors.
  geom_density(color = "red", size = 1) + #And we make a density plot.
  labs(title = "Log transformed HS grad", x = "Transformed HS grad", y = "Density") #We make the titles and labels.

ggplot(df_states, aes(x = sqrt_HS_grad)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Sqrt transformed HS grad", x = "Transformed HS grad", y = "Density")
```

It still doesn't really look normal, but let's check QQ-plots.

```{r}
#Now for the QQ-Plots
ggplot(df_states, aes(sample = log_HS_grad)) + #He start the plot somewhat the same way as before.
  stat_qq(color = "blue") + # We do the colors
  stat_qq_line(color = "red") + #And we make the QQ line
  labs(title = "Q-Q Plot: log of HS grad")

ggplot(df_states, aes(sample = sqrt_HS_grad)) + #He start the plot somewhat the same way as before.
  stat_qq(color = "lightgreen") + # We do the colors
  stat_qq_line(color = "black") + #And we make the QQ line
  labs(title = "Q-Q Plot: sqrt of HS grad")
```

Maybe it looks a little better than before, but let's check withg Shapiro-Wilk test

```{r}
shapiro.test(df_states$log_HS_grad)
shapiro.test(df_states$sqrt_HS_grad)
```
So we cannot fix the normality problems of the HS grad, let's try the murder one.

```{r}

# Log transformation
df_states$log_murder <- log(df_states$Murder)

# Sqrt transformation
df_states$sqrt_murder <- sqrt(df_states$Murder)


ggplot(df_states, aes(x = log_murder)) + #Here, we specify that we look at HS grad on the x-axis.
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue", color = "black") + #And here we make the histogram, and specify the bin length and colors.
  geom_density(color = "red", size = 1) + #And we make a density plot.
  labs(title = "Log transformed murder", x = "Transformed murder", y = "Density") #We make the titles and labels.

ggplot(df_states, aes(x = sqrt_murder)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Sqrt transformed murder", x = "Transformed murder", y = "Density")
```

Okay, let's see the QQ-plots
```{r}
#Now for the QQ-Plots
ggplot(df_states, aes(sample = log_murder)) + #He start the plot somewhat the same way as before.
  stat_qq(color = "blue") + # We do the colors
  stat_qq_line(color = "red") + #And we make the QQ line
  labs(title = "Q-Q Plot: log of murder")

ggplot(df_states, aes(sample = sqrt_murder)) + #He start the plot somewhat the same way as before.
  stat_qq(color = "lightgreen") + # We do the colors
  stat_qq_line(color = "black") + #And we make the QQ line
  labs(title = "Q-Q Plot: sqrt of murder")
```
Maybe it's a little better it seems, but let's do the test.

```{r}
shapiro.test(df_states$log_murder)
shapiro.test(df_states$sqrt_murder)
```

The p-value of the log transformed murder, was still significant, but the sqrt transformation worked and it's now normal. But since HS grad couldn't become normal, we still do the spearman test:

```{r}
cor.test(df_states$Murder, df_states$`HS Grad`, method = "spearman")
```
The sparmans correlational test found that the high school graduation rate and the number of murders per capita was found to be significantly correlated r(48) = -0.44, p = 0.002.

This means that they are negatively correlated, meaning that the more murders, the lower the highschool graduation rate and vice versa.
